# Data Processing {#sec-data-processing}

```{r}
#| label: setup
#| include: false

library(here)
library(geodata)
library(leaflet)
library(sf)

here("R", ".setup.R") |> source()
```

::: {.callout-note}
`LogoClim` integrates [WorldClim](https://www.worldclim.org/) data into [NetLogo](https://www.netlogo.org/) models through a multi-step conversion process that can introduce minor discrepancies between the original data and the values used in the model. We recommend verifying data accuracy by running the [Near-Equality Tests](#sec-sm-near-equality-tests) notebook before proceeding with your analysis.
:::

To convert the downloaded [GeoTIFF](https://en.wikipedia.org/wiki/GeoTIFF) files to [Esri ASCII](https://en.wikipedia.org/wiki/Esri_grid) grid format, we developed the [`worldclim_to_ascii()`](https://danielvartan.github.io/orbis/reference/worldclim_to_ascii.html) function, also part of the [`orbis`](https://danielvartan.github.io/orbis/index.html) R package [@vartanian2026a]. Just like [`worldclim_download()`](https://danielvartan.github.io/orbis/reference/worldclim_download.html), it is pretty straightforward to use. Here is a overview of its main parameters:

```r
worldclim_to_ascii(
  file,
  shape = NULL
  box = NULL
)
```

[`worldclim_to_ascii`](https://danielvartan.github.io/orbis/reference/worldclim_to_ascii.html) allow you not only convert the [GeoTIFF](https://en.wikipedia.org/wiki/GeoTIFF) files to [Esri ASCII](https://en.wikipedia.org/wiki/Esri_grid) format, but also crop the data to a specific region of interest. This is done using the `shape` or `box` parameters.

The first parameter, `shape`, expects a [spatial object](https://rspatial.github.io/terra/reference/SpatRaster-class.html) that defines the area to crop. The second parameter, `box`, allows you to specify a [bounding box](https://wiki.openstreetmap.org/wiki/Bounding_box) using numeric coordinates. You can use one or the other, but not both at the same time. If neither is provided, the function will convert the entire dataset without cropping.

## Shapes

To obtain shapes of countries and their administrative boundaries, we recommend using the [`geodata`](https://rspatial.github.io/geodata/) R package [@hijmans2024], which provides access to shapes from the [GADM](https://gadm.org) database [@hijmansa]. You can install in [R](https://www.r-project.org/) with:

```r
install.packages("geodata")
```

And load it with:

```{r}
#| output: false

library(geodata)
```

With `geodata` loaded, use the [`gadm()`](https://rspatial.github.io/geodata/reference/gadm.html) function to download country shapes. This function identifies countries using the [ISO 3166-1 alpha-3](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-3) standard and supports multiple resolutions, from country boundaries (level 0) to administrative subdivisions (levels 1, 2, and 3).

For example, if you want to crop [WorldClim](https://www.worldclim.org/) data to the boundaries of Brazil, you could use the following code:

```{r}
#| label: Download country shape
#| output: false

country_shape <- gadm("BRA", path = "data", level = 0)
```

This code will download the shape of Brazil and store it in the `country_shape` variable, which can then be passed to the `shape` parameter of the [`worldclim_to_ascii`](https://danielvartan.github.io/orbis/reference/worldclim_to_ascii.html) function for cropping.

```{r}
#| label: Plot country shape
#| code-fold: true

library(leaflet)
library(sf)

leaflet() |>
  addProviderTiles(providers$Esri.WorldStreetMap) |>
  fitBounds(
    lng1 = country_shape |>
      st_bbox() |>
      magrittr::extract("xmin") |>
      unname(),
    lat1 = country_shape |>
      st_bbox() |>
      magrittr::extract("ymin") |>
      unname(),
    lng2 = country_shape |>
      st_bbox() |>
      magrittr::extract("xmax") |>
      unname(),
    lat2 = country_shape |>
      st_bbox() |>
      magrittr::extract("ymax") |>
      unname()
  ) |>
  addPolygons(
    data = country_shape,
    fillColor = "transparent",
    color = "blue",
    weight = 2,
    opacity = 1
  )
```

## Boxes

Alternatively, you can specify a [bounding box](https://wiki.openstreetmap.org/wiki/Bounding_box) using the `box` parameter. This parameter expects a numeric vector with four values: `c(xmin, ymin, xmax, ymax)`, where `xmin` and `xmax` are the minimum and maximum longitudes, and `ymin` and `ymax` are the minimum and maximum latitudes of the area you want to crop.

For example, if you want to crop the entire European region, you could use the following code:

```{r}
box <- c(-24.534, 34.625, 50.601, 71.181)
```

```{r}
#| label: Plot bounding box
#| code-fold: true

library(leaflet)

leaflet() |>
  addProviderTiles(providers$Esri.WorldStreetMap) |>
  addRectangles(
    lng1 = box[1],
    lat1 = box[2],
    lng2 = box[3],
    lat2 = box[4],
    fillColor = "transparent",
    color = "blue",
    weight = 2,
    opacity = 1
  )
```

::: {.callout-tip}
Use the [`bbox finder`](https://bboxfinder.com) tool to easily obtain the coordinates for your area of interest.
:::

## Conversion

With the files downloaded and the shape or box defined, you can proceed to convert the data using the [`worldclim_to_ascii`](https://danielvartan.github.io/orbis/reference/worldclim_to_ascii.html) function. This function will take care of the conversion and cropping (if specified) in one step.

For example, to work with average maximum temperature from the *Historical Climate Data* series at 10-minute resolution for Brazil, you could use the following code:

```r
library(orbis)
```

```r
files <- worldclim_download(
  series = "hcd",
  resolution = "10m",
  variable = "tmax",
  dir = "path/to/your/data/directory"
)
#> ✔ Scraping WorldClim website [688ms]
#> ℹ Total download size (compressed): 35.7M.
#> ✔ Calculating file sizes [614ms]
#> ✔ Creating LICENSE and README files [291ms]
#> ℹ Downloading 1 file to /path/to/your/data/directory/historical-climate-data
#> ✔ Downloading files [37.3s]
#> ✔ Unzipping files [14ms]
```

```r
country_shape <- gadm("BRA", level = 0)
```

```r
files |> worldclim_to_ascii(shape = country_shape)
#> Converting data ■■■■■■■■■■■■■■■■■■■■■■■■■■■■■■■  100% | ETA:  0s
```

The function default values already takes care of possible distortions caused by geographic projection or rotating issues (e.g., shapes that cross the [International Date Line](https://en.wikipedia.org/wiki/International_Date_Line)). It also removes abnormal outliers from the data.

After processing the data, you're ready to use `LogoClim`. Just open the model and point at your data directory using the model interface.

For more details on available conversion parameters, refer to the [`worldclim_to_ascii`](https://danielvartan.github.io/orbis/reference/worldclim_to_ascii.html) documentation.
