---
execute:
  eval: false
---

# WorldClim 2.1: Data Download

## Overview

This document provides a step-by-step guide to download the WorldClim 2.1 data from the [WorldClim website](https://worldclim.org/data/monthlywth.html).

::: {.callout-important}
This process may take some time to complete. Please be patient.
:::

## Setting the Environment

```{r}
#| eval: false

library(beepr)
library(checkmate)
library(cli)
library(curl)
library(dplyr)
library(fs)
library(here)
library(httr)
library(magrittr)
library(purrr)
library(rutils) # github.com/danielvartan/rutils (GitHub)
library(rvest)
library(stringr)
library(utils)
library(zip)
```

```{r}
source(here::here("R", "download_wc_files.R"))
source(here::here("R", "get_content_length.R"))
source(here::here("R", "zip_wc_files.R"))
```

## Setting Variables

### Options

```{r}
options(cli.progress_show_after = 0)
```

### Series

```{r}
series <- "historical-climate-data"
source <- "https://worldclim.org/data/worldclim21.html"
```

```{r}
#| eval: false

series <- "historical-monthly-weather-data"
source <- "https://worldclim.org/data/monthlywth.html"
```

```{r}
#| eval: false

series <- "future-climate-data"

# # Future climate, 10 minutes spatial resolution
# source <- "https://worldclim.org/data/cmip6/cmip6_clim10m.html"
# Future climate, 5 minutes spatial resolution
source <- "https://worldclim.org/data/cmip6/cmip6_clim5m.html"
# # Future climate, 2.5 minutes spatial resolution
# source <- "https://worldclim.org/data/cmip6/cmip6_clim2.5m.html"
# # Future climate, 30 seconds spatial resolution
# source <- "https://worldclim.org/data/cmip6/cmip6_clim30s.html"
```

### Directories

```{r}
raw_data_dir <- here::here("data-raw")
```

```{r}
wc_raw_data_dir <- fs::path(raw_data_dir, "worldclim")
```

```{r}
wc_raw_data_series_dir <- fs::path(wc_raw_data_dir, series)
```

```{r}
dirs <- c(raw_data_dir, wc_raw_data_dir, wc_raw_data_series_dir)

for (i in dirs) {
  if (!checkmate::test_directory_exists(i)) {
    fs::dir_create(i, recurse = TRUE)
  }
}
```

### Parameters

```{r}
resolution <- "5m" # all, 10m, 5m, 2.5m, 30s
```

```{r}
# This is only needed for future climate data.
model <- NULL # "ACCESS-CM2"
```

## Scrapping Source

```{r}
html <- source |> rvest::read_html()
```

## Extracting URLs

```{r}
urls <- html |>
  rvest::html_elements("a") |>
  rvest::html_attr("href") |>
  stringr::str_subset("geodata")
```

```{r}
if (!resolution == "all") {
  urls <-
    urls %>%
    magrittr::extract(stringr::str_detect(basename(.), resolution))
}
```

## Creating the Metadata

```{r}
sizes <-
  urls |>
  purrr::map_dbl(
    .f = get_content_length,
    .progress = TRUE
)

beepr::beep(1)
```

```{r}
metadata <- dplyr::tibble(
  url = urls,
  file = basename(urls),
  size = as.numeric(sizes), # In bytes # 1 megabyte = 2^20 bytes
  size_cum_sum = cumsum(dplyr::coalesce(size, 0)) + (size * 0)
)
```

```{r}
metadata |> dplyr::glimpse()
```

```{r}
models <-
  metadata |>
  magrittr::extract2("file") |>
  stringr::str_extract("(?<=(bioc_|tmin_|tmax_|prec_)).*(?=_ssp)") |>
  unique()

models
```

## Checking for Errors

```{r}
{
  cli::cli_alert_info(
    paste0(
      "{.strong {cli::col_red(rutils:::count_na(metadata$size))}} ",
      "url requests resulted in error."
    )
  )

  if (rutils::count_na(metadata$size) > 0) {
    cli::cli_alert_info("Their file names are:")
    cli::cli_li(metadata$file[is.na(metadata$size)])
  }
}
```

## Downloading Files

```{r}
#| eval: false

zip_dir <- fs::path(wc_raw_data_series_dir, "zip")

if (!checkmate::test_directory_exists(zip_dir)) {
  zip_dir |> fs::dir_create(recurse = TRUE)
}
```

```{r}
broken_links <-
  metadata |>
  download_wc_files(dir = zip_dir, model = model)

beepr::beep(1)
```

```{r}
#| eval: false

# Use only if the download process was interrupted.

# downloaded_files <-
#   file.path(wc_raw_data_dir, "zip") |>
#   fs::dir_ls(type = "file")

# broken_links <-
#   metadata |>
#   # dplyr::filter(stringr::str_detect(file, model)) |>
#   dplyr::filter(!file %in% downloaded_files) |>
#   download_wc_files(dir = file.path(wc_raw_data_dir, "zip"))

# beepr::beep(1)
```

## Compressing Files (Just for Future Climate Data)

```{r}
#| eval: false
#| output: false

suffix <- paste0("_", "2-5m")

metadata |>
  zip_wc_files(
    model = model,
    dir = dir,
    suffix = suffix,
    broken_links = broken_links,
    engine = "utils" # "zip"
  )

beepr::beep(1)
```

## Deleting Downloaded Files (Just for Future Climate Data)

```{r}
#| eval: false

# Careful! This routine should be use only in special cases.

for (i in model) {
  list.files(raw_data_dir, full.names = TRUE) |>
    stringr::str_subset(i) |>
    file.remove()
}

invisible(gc())
```
