# Procedure Behavior Tests {#sec-sm-procedure-behavior-tests}

```{r}
#| label: Setup
#| include: false

library(here)

here("R", ".setup.R") |> source()
```

This document provides unit tests for the [`LogoClim`](https://github.com/sustentarea/logoclim) NetLogo model. The tests validate [procedure](https://docs.netlogo.org/programming.html#procedures) behavior to ensure correct functionality.

## Problem

NetLogo [procedures](https://docs.netlogo.org/programming.html#procedures) can produce unexpected results and exhibit side effects that may go undetected during development. The unit tests in this document validate procedure behavior through [expectation tests](https://testthat.r-lib.org/reference/equality-expectations.html), ensuring they function as intended and produce reliable results.

## Methods

### Data munging

All processes were made using the [Quarto](https://quarto.org/) publishing system [@allaire], the [NetLogo](https://www.netlogo.org/) environment, the [R](https://www.r-project.org/) programming language [@rcoreteam], and several R packages.

For data manipulation and workflow, priority was given to packages from the [Tidyverse](https://www.tidyverse.org/), [rOpenSci](https://ropensci.org/) and [rspatial](https://rspatial.org) ecosystems, as well as other packages adhering to the tidy tools manifesto [@wickham2023c].

### NetLogo Integration

Integration with [NetLogo](https://www.netlogo.org/) [@wilensky1999a] is facilitated by the [`logolink`](https://danielvartan.github.io/logolink/) R package [@vartanian2026]. This package enables the execution of [BehaviorSpace](https://docs.netlogo.org/behaviorspace.html) experiments directly from R.

Output is extracted in [Table](https://docs.netlogo.org/behaviorspace.html#table-output) or [Lists](https://docs.netlogo.org/behaviorspace.html#lists-output) format, depending on the experiment configuration.

No [Java](https://www.java.com/en/) dependencies are required. NetLogo bundles its own Java Runtime Environment (JRE), ensuring independent operation regardless of the system's Java installation.

### Continuous Integration

These tests use the latest release of [NetLogo](https://www.netlogo.org/) and are automated using [GitHub Actions](https://github.com/features/actions) provided by the [`LogoActions`](https://github.com/danielvartan/logoactions) project [@vartanian2026b]. Each commit to the code repository triggers test execution, ensuring that errors are caught early in the development process.

### Expectation Tests

Expectations are validated using the [`testthat`](https://testthat.r-lib.org/) [@wickham2011] and [`checkmate`](https://mllg.github.io/checkmate/) [@lang2017] R packages.

### Code Style

The Tidyverse [Tidy Tools Manifesto](https://tidyverse.tidyverse.org/articles/manifesto.html) [@wickham2023c], [code style guide](https://style.tidyverse.org/) [@wickhama] and [design principles](https://design.tidyverse.org/) [@wickhamc] were followed to ensure consistency and enhance readability.

### Reproducibility

The pipeline is fully reproducible and can be run again at any time. To ensure consistent results, the [`renv`](https://rstudio.github.io/renv/) package [@ushey2025] was used to manage and restore the R environment. See the [README](https://github.com/sustentarea/logoclim/blob/main/README.md) file in the code repository to learn how to run it.

## Set the Environment

### Load Packages

```{r}
#| label: Set the Environment
#| output: false

library(cli)
library(dplyr)
library(here)
library(logolink)
library(magrittr)
library(purrr)
library(testthat)
```

### Set Initial Variables

::: {.callout-note}
Setting the `JAVA_TOOL_OPTIONS` is optional, but recommended to avoid unnecessary messages from the Java Media Framework.
:::

```{r}
Sys.setenv(JAVA_TOOL_OPTIONS = "-Dcom.sun.media.jai.disableMediaLib=true")
```

```{r}
model_path <- here("nlogox", "logoclim.nlogox")
```

## `as-list`

```{r}
#| label: as-list

setup_file <- create_experiment(
  name = "as-list",
  setup = "setup false",
  go = NULL,
  metrics = c(
    'as-list "a"',
    'as-list 1',
    'as-list true',
    'as-list [1 2 3]'
  )
)
```

```{r}
results <-
  model_path |>
  run_experiment(
    setup_file = setup_file
  )
```

```{r}
results |>
  pluck("table") |>
  glimpse()
```

```{r}
test_that("as-list", {
  results |>
    extract2("table") |>
    pull(as_list_a) |>
    expect_equal("[a]")

  results |>
    extract2("table") |>
    pull(as_list_1) |>
    expect_equal("[1]")

  results |>
    extract2("table") |>
    pull(as_list_true) |>
    expect_equal("[true]")

  results |>
    extract2("table") |>
    pull(as_list_1_2_3) |>
    expect_equal("[1 2 3]")
})
```

## `quartile`

```{r}
#| label: quartile

setup_file <- create_experiment(
  name = "quartile",
  setup = "setup false",
  go = NULL,
  metrics = c(
    'quartile [0 1 2 3 4] 0',
    'quartile [0 1 2 3 4] 1',
    'quartile [0 1 2 3 4] 2',
    'quartile [0 1 2 3 4] 3',
    'quartile [0 1 2 3 4] 4',
    'quartile [0 1 2 3 4] "iqr"',
    'quartile [0 1 2 3 4] "length"',
    'quartile [0 1 2 3 4 5] 0',
    'quartile [0 1 2 3 4 5] 1',
    'quartile [0 1 2 3 4 5] 2',
    'quartile [0 1 2 3 4 5] 3',
    'quartile [0 1 2 3 4 5] 4',
    'quartile [0 1 2 3 4 5] "iqr"',
    'quartile [0 1 2 3 4 5] "length"'
  )
)
```

```{r}
results <-
  model_path |>
  run_experiment(
    setup_file = setup_file
  )
```

```{r}
results |>
  pluck("table") |>
  glimpse()
```

```{r}
test_that("quartile [0 1 2 3 4]", {
  results |>
    extract2("table") |>
    pull(quartile_0_1_2_3_4_0) |>
    expect_equal(0)

  results |>
    extract2("table") |>
    pull(quartile_0_1_2_3_4_1) |>
    expect_equal(0)

  results |>
    extract2("table") |>
    pull(quartile_0_1_2_3_4_2) |>
    expect_equal(1)

  results |>
    extract2("table") |>
    pull(quartile_0_1_2_3_4_3) |>
    expect_equal(2)

  results |>
    extract2("table") |>
    pull(quartile_0_1_2_3_4_4) |>
    expect_equal(4)

  results |>
    extract2("table") |>
    pull(quartile_0_1_2_3_4_iqr) |>
    expect_equal(2 - 0)

  results |>
    extract2("table") |>
    pull(quartile_0_1_2_3_4_length) |>
    expect_equal(5 / 4)
})
```

```{r}
test_that("quartile [0 1 2 3 4 5]", {
  results |>
    extract2("table") |>
    pull(quartile_0_1_2_3_4_5_0) |>
    expect_equal(0)

  results |>
    extract2("table") |>
    pull(quartile_0_1_2_3_4_5_1) |>
    expect_equal(0)

  results |>
    extract2("table") |>
    pull(quartile_0_1_2_3_4_5_2) |>
    expect_equal(2)

  results |>
    extract2("table") |>
    pull(quartile_0_1_2_3_4_5_3) |>
    expect_equal(3)

  results |>
    extract2("table") |>
    pull(quartile_0_1_2_3_4_5_4) |>
    expect_equal(5)

  results |>
    extract2("table") |>
    pull(quartile_0_1_2_3_4_5_iqr) |>
    expect_equal(3 - 0)

  results |>
    extract2("table") |>
    pull(quartile_0_1_2_3_4_5_length) |>
    expect_equal(6 / 4)
})
```

## `single-quote`

```{r}
#| label: single-quote

setup_file <- create_experiment(
  name = "single-quote",
  setup = "setup false",
  go = NULL,
  metrics = c(
    'single-quote 1',
    'single-quote [1 2 3]'
  )
)
```

```{r}
results <-
  model_path |>
  run_experiment(
    setup_file = setup_file
  )
```

```{r}
results |>
  pluck("table") |>
  glimpse()
```

```{r}
test_that("single-quote", {
  results |>
    extract2("table") |>
    pull(single_quote_1) |>
    expect_equal("'1'")

  results |>
    extract2("table") |>
    pull(single_quote_1_2_3) |>
    expect_equal("['1' '2' '3']")
})
```

## `str-to-num-month`

```{r}
#| label: str-to-num-month

setup_file <- create_experiment(
  name = "str-to-num-month",
  setup = "setup false",
  go = NULL,
  metrics = c(
    'str-to-num-month "January"',
    'str-to-num-month "July"',
    'str-to-num-month "December"'
  )
)
```

```{r}
results <-
  model_path |>
  run_experiment(
    setup_file = setup_file
  )
```

```{r}
results |>
  pluck("table") |>
  glimpse()
```

```{r}
test_that("str-to-num-month", {
  results |>
    extract2("table") |>
    pull(str_to_num_month_january) |>
    expect_equal(1)

  results |>
    extract2("table") |>
    pull(str_to_num_month_july) |>
    expect_equal(7)

  results |>
    extract2("table") |>
    pull(str_to_num_month_december) |>
    expect_equal(12)
})
```

## `unique-outliers`

```{r}
#| label: unique-outliers

setup_file <- create_experiment(
  name = "unique-outliers",
  setup = "setup false",
  go = NULL,
  metrics = c(
    'unique-outliers [1 2 3 4 100 100 500 500] 1.5',
    'unique-outliers [1 2 3] 3'
  )
)
```

```{r}
results <-
  model_path |>
  run_experiment(
    setup_file = setup_file
  )
```

```{r}
results |>
  pluck("table") |>
  glimpse()
```

```{r}
test_that("unique-outliers", {
  results |>
    extract2("table") |>
    pull(unique_outliers_1_2_3_4_100_100_500_500_1_5) |>
    expect_equal("[100 500]")

  results |>
    extract2("table") |>
    pull(unique_outliers_1_2_3_3) |>
    expect_equal("[]")
})
```
