# Procedure Behavior Tests {#sec-sm-procedure-behavior-tests}

```{r}
#| label: Setup
#| include: false

library(here)

here("R", ".setup.R") |> source()
```

This document provides unit tests for the [`LogoClim`](https://github.com/sustentarea/logoclim) NetLogo model. The tests validate [procedure](https://docs.netlogo.org/programming.html#procedures) behavior and arguments to ensure correct model functionality.

## Problem

[Procedures](https://docs.netlogo.org/programming.html#procedures) can produce unexpected results and have side effects that may not be captured during the development process. This document addresses these challenges by providing pre-defined expectations for procedure behaviors.

## Methods

### NetLogo Integration

Integration with [NetLogo](https://www.netlogo.org/) [@wilensky1999a] is facilitated by the [`logolink`](https://danielvartan.github.io/logolink/) R package [@vartanian2026]. This package enables the execution of [BehaviorSpace](https://docs.netlogo.org/behaviorspace.html) experiments directly from R.

Outputs were extracted in [Table](https://docs.netlogo.org/behaviorspace.html#table-output) and [Lists](https://docs.netlogo.org/behaviorspace.html#lists-output) format, along with metadata about the model's settings.

No [Java](https://www.java.com/en/) dependencies are required. NetLogo bundles its own Java Runtime Environment (JRE), ensuring independent operation regardless of the system's Java installation.

### Continuous Integration

These tests use the latest release of [NetLogo](https://www.netlogo.org/) and are automated using [GitHub Actions](https://github.com/features/actions) provided by the [`LogoActions`](https://github.com/danielvartan/logoactions) project [@vartanian2026b]. Each commit to the code repository triggers test execution, ensuring that errors are caught early in the development process.

### Expectation Tests

Expectations are validated using the [`testthat`](https://testthat.r-lib.org/) [@wickham2011] and [`checkmate`](https://mllg.github.io/checkmate/) [@lang2017] R packages.

### Code Style

The Tidyverse [Tidy Tools Manifesto](https://tidyverse.tidyverse.org/articles/manifesto.html) [@wickham2023c], [code style guide](https://style.tidyverse.org/) [@wickhama] and [design principles](https://design.tidyverse.org/) [@wickhamc] were followed to ensure consistency and enhance readability.

### Reproducibility

The pipeline is fully reproducible and can be run again at any time. To ensure consistent results, the [`renv`](https://rstudio.github.io/renv/) package [@ushey2025] was used to manage and restore the R environment. See the [README](https://github.com/sustentarea/logoclim/blob/main/README.md) file in the code repository to learn how to run it.

## Set the Environment

### Load Packages

```{r}
#| label: Set the Environment

library(cli)
library(dplyr)
library(here)
library(logolink)
library(magrittr)
library(testthat)
```

### Set Initial Variables

::: {.callout-note}
Setting the `JAVA_TOOL_OPTIONS` is optional, but recommended to avoid unnecessary messages from the Java Media Framework.
:::

```{r}
Sys.setenv(JAVA_TOOL_OPTIONS = "-Dcom.sun.media.jai.disableMediaLib=true")
```

```{r}
model_path <- here("nlogox", "logoclim.nlogox")
```

## `single-quote`

```{r}
#| label: single-quote

setup_file <- create_experiment(
  name = "single-quote",
  setup = "setup false",
  metrics = c(
    'single-quote 1',
    'single-quote [1 2 3]'
  )
)
```

```{r}
results <-
  model_path |>
  run_experiment(
    setup_file = setup_file
  )
```

```{r}
results |>
  extract2("table") |>
  pull(single_quote_1) |>
  expect_equal("'1'")
```

```{r}
results |>
  extract2("table") |>
  pull(single_quote_1_2_3) |>
  expect_equal("['1' '2' '3']")
```
