# Welcome {.unnumbered}

```{r}
#| label: setup
#| include: false

library(badger)
library(here)
library(htmltools)
library(knitr)

here("R", ".setup.R") |> source()
```

::: {.callout-important}
This manual is still under development and may be subject to change.

This warning will be removed once the manual is finalized.
:::

This manual provides a comprehensive overview of the [`LogoClim`](https://github.com/sustentarea/logoclim) NetLogo model. The documentation includes detailed descriptions of the model's components, usage instructions, and examples to help users get started.

If you feel you still need more help, please open a discussion in the model's [Discussion tab](https://github.com/danielvartan/orbis/discussions) on GitHub. While we can't guarantee immediate assistance, the community is here to help.

## Citation

```{r}
#| echo: false
#| output: asis

cat(
  badge_custom(
    x = "doi",
    y = "10.17605/OSF.IO/EAPZU",
    color = "1284C5",
    url = "https://doi.org/10.17605/OSF.IO/EAPZU",
    alt = ""
  )
)
```

::: {.callout-important}
When using WorldClim data, you must also cite the original data sources. The appropriate citation depends on the specific dataset utilized. Please refer to the [WorldClim website](https://www.worldclim.org/data/index.html#citation) for up-to-date citation guidelines and dataset references.
:::

If you use this model in your research, please cite it to acknowledge the effort invested in its development and maintenance. Your citation helps support the ongoing improvement of the model.

To cite `LogoClim` in publications please use the following format:

Vartanian, D., Garcia, L., & Carvalho, A. M. (2026). *LogoClim: WorldClim in NetLogo* [Computer software]. <https://doi.org/10.17605/OSF.IO/EAPZU>

A BibTeX entry for LaTeX users is:

```latex
@Misc{vartanian2026,
  title = {LogoClim: WorldClim in NetLogo},
  author = {{Daniel Vartanian} and {Leandro Garcia} and {Aline Martins de Carvalho}},
  year = {2026},
  doi = {10.17605/OSF.IO/EAPZU},
  note = {Computer software}
}
```

## Contributing

```{r}
#| echo: false
#| output: asis

cat(
  badge_custom(
    x = "Contributor%20Covenant",
    y = "3.0",
    color = "4baaaa",
    url = "https://www.contributor-covenant.org/version/3/0/code_of_conduct/",
    alt = ""
  )
)
```

Contributions are always welcome! Whether you want to report bugs, suggest new features, or help improve the code or documentation (including this manual), your input makes a difference.

Before opening a new issue, please check the model's [issues tab](https://github.com/sustentarea/logoclim/issues) to see if your topic has already been reported.

[![](https://img.shields.io/static/v1?label=Sponsor&message=%E2%9D%A4&logo=GitHub&color=%23fe8e86)](https://github.com/sponsors/danielvartan)

You can also support the development of `LogoClim` by becoming a sponsor.

Click [here](https://github.com/sponsors/danielvartan) to make a donation. Please mention `LogoClim` in your donation message.

## License

```{r}
#| echo: false
#| output: asis

cat(
  badge_license(
    license = "CC BY-NC-SA",
    color = "lightgrey",
    url = "https://creativecommons.org/licenses/by-nc-sa/4.0/",
    alt = ""
  )
)
```

This manual is licensed under the [Creative Commons Attribution-NonCommercial-ShareAlike 4.0 International License](https://creativecommons.org/licenses/by-nc-sa/4.0/). You can share and adapt the material for non-commercial purposes, as long as you provide appropriate credit, link to the license, and indicate if changes were made.

## Acknowledgments

We gratefully acknowledge [Robert J. Hijmans](https://orcid.org/0000-0001-5872-2872), [Stephen E. Fick](https://orcid.org/0000-0002-3548-6966), and the entire [WorldClim](https://worldclim.org/) team for their outstanding work in creating and maintaining the WorldClim datasets.

We thank the [Climatic Research Unit](https://www.uea.ac.uk/groups-and-centres/climatic-research-unit) at the [University of East Anglia](https://www.uea.ac.uk/) and the United Kingdom's [Met Office](https://www.metoffice.gov.uk/) for developing and providing access to the [CRU-TS-4.09](https://crudata.uea.ac.uk/cru/data/hrg/cru_ts_4.09/) dataset, a vital source of historical climate data.

We also acknowledge the World Climate Research Programme ([WCRP](https://www.wcrp-climate.org/)), its Working Group on Coupled Modelling, and the Coupled Model Intercomparison Project Phase 6 ([CMIP6](https://pcmdi.llnl.gov/CMIP6/)) for coordinating and advancing global climate model development.

We are grateful to the climate modeling groups for producing and sharing their model outputs, the Earth System Grid Federation ([ESGF](https://esgf.llnl.gov/)) for archiving and providing access to the data, and the many funding agencies that support [CMIP6](https://pcmdi.llnl.gov/CMIP6/) and [ESGF](https://esgf.llnl.gov/).

```{r}
#| echo: false
#| output: asis

tags$table(
  tags$tr(
    tags$td(
      tags$p(
        tags$a(
          tags$img(
            # Don't use `here()` here.
            src = "images/sustentarea-logo.svg",
            width = 115
          ),
          href = "https://www.fsp.usp.br/sustentarea/"
        ),
        align = "center"
      ),
      width = "30%",
      valign = "middle"
    ),
    tags$td(
      tags$p(
        "This work was developed with support from the ",
        tags$a(
          "Sustentarea",
          href = "https://www.fsp.usp.br/sustentarea/"
        ),
        " Research and Extension Center at the University of São Paulo (",
        tags$a(
          "USP",
          href = "https://www5.usp.br/",
          .noWS = "outside"
        ),
        ")."
      ),
      width = "70%",
      valign = "middle"
    )
  )
) |>
  raw_html()

tags$table(
  tags$tr(
    tags$td(
      tags$p(
        tags$a(
          tags$img(
            # Don't use `here()` here.
            src = "images/resiclima-logo.svg",
            width = 115
          ),
          href = "https://resiclima.com.br/"
        ),
        align = "center"
      ),
      width = "30%",
      valign = "middle"
    ),
    tags$td(
      tags$p(
        "This work was developed with support from the ",
        tags$a(
          "Resiclima Network",
          href = "https://resiclima.com.br/",
          .noWS = "outside"
        ),
        ", an international collaboration for the multidimensional and ",
        "interdisciplinary study of global climate change."
      ),
      width = "70%",
      valign = "middle"
    )
  )
) |>
  raw_html()

tags$table(
  tags$tr(
    tags$td(
      tags$p(
        tags$a(
          tags$img(
            # Don't use `here()` here.
            src = "images/cnpq-logo.svg",
            width = 150
          ),
          href = "https://www.gov.br/cnpq/"
        ),
        align = "center"
      ),
      width = "30%",
      valign = "middle"
    ),
    tags$td(
      tags$p(
        "This work was supported by the Department of Science and ",
        "Technology of the Secretariat of Science, Technology, and Innovation ",
        "and of the Health Economic-Industrial Complex (",
        tags$a(
          "SECTICS",
          href = "https://www.gov.br/saude/pt-br/composicao/sectics/",
          .noWS = "outside"
        ),
        ")  of the ",
        tags$a(
          "Ministry of Health",
          href = "https://www.gov.br/saude/pt-br/composicao/sectics/",
          .noWS = "outside"
        ),
        " ",
        "of Brazil, and the National Council for Scientific and ",
        "Technological Development (",
        tags$a(
          "CNPq",
          href = "https://www.gov.br/cnpq/",
          .noWS = "outside"
        ),
        ") (grant no. 444588/2023-0)."
      ),
      width = "70%",
      valign = "middle"
    )
  )
) |>
  raw_html()
```
